Composition-climate paper

### 2016-11-11 Composition_Ecology.Rmd
All changes form *historic* to *historical*.

# Effect of historic land-use and climate change on tree-climate relationships in the northern United States

@@

Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions is poorly understood. Here we show that tree-climate relationships for 15 common tree genera in the northern US have significantly altered over the last two centuries due to historic land use, pathogen outbreaks, and climate change. Realized niches have shifted towards higher minimum temperatures, cooler maximum temperatures, and higher rainfall. A new attribution method implicates both historic climate change and land use in these shifts, with the relative importance varying among genera and climate variables. Most climate/land-use interactions are compounding, in which historic land-use reinforces climatic shifts in species distributions toward cooler or wetter distributions. The prevalence of compounding interactions imply that contemporary-based models of species distributions may underestimate species resilience to climate change.

@@

Historic land use has substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree] and may cause species distributions to incompletely fill available climates, or otherwise be at disequilibrium [@early2014climatic;@sarmento2013impacts;@ivory2016niche]. Anthropogenic land use is structured and is often biased within climate space, hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature]. Biasing can confound studies that use contemporary distributional data to understand the environmental controls on species distributions and predict species responses to future climate scenarios [@pyke2004habitat].

@@

Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and we diagnose the relative influence of historic climate and land use changes on these shifts.  We use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.  We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.  We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and attribute the relative importance of historic vegetation and climate change. We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts. We use this framework to discuss the potential impacts of historical land-use and climate change on inferences about species resilience to climate change, as predicted by species-distribution models based soley on modern distributional data.

@@

`r fig_nums("fourPanel", "(a) The 2x2 factorial design used to diagnose the effects of historic climate and vegetation change on observed shifts in tree-climate relationships. Observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row). (b-e) Representative normalized probability densities for *Larix* for each of the four combinations of climate and vegetation states, plotted relative to $T_{max}$ (b), $T_{diff}$ (c), $T_{min}$ (d), and $P_{ann}$ (e).")`

@@

For most tree genera, historic range losses dominate over range gains and are spatially structured (Table 1, Figure 3). Range losses in the upper Midwest are concentrated in the south and west, along the prairie-savanna-forest ecotone (Figure 3). For *Ulmus* and *Larix*, however, losses occur throughout the range.  Losses are highest for *Ostrya*/*Carpinus* and *Larix* when assessed as a proportion of total range (Supplemental Table 1), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the Upper Midwest.  Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed for *Populus* (`r fig_nums("gain_loss", display="cite")`) and averages `r round(mean(loss_table$Gain),1)`% across all taxa.   Spatial patterns of range gains vary widely among tree genera, and often show less spatial bias than losses (Figure 3).  Range gains for *Picea*, *Larix*, *Abies*, *Tsuga*, *Fagus*, *Ostrya*/*Carpinus*, *Tilia*, and *Fraxinus*, for example, all show a dispersed pattern of gains in a relatively few and widely scattered cells, suggesting that these apparent range gains are mainly caused by sampling uncertainty inherent in local-scale plots and stand heterogeneity, rather than systematic range gains.  Conversely, gains for *Populus* and *Acer*, are concentrated in northern Minnesota, Wisconsin, and Michigan, suggesting range infilling, while *Quercus* gains are consistent with some range extension northwards.  *Pinus*, *Juniperus*/*Thuja*, and *Ulmus* range gains are limited but primarily in the southern portion of their historic range, or beyond it.

@@

The joint effects of historic climate change and land use on tree-climate distributions are illustrated for *Larix* (Figure 2).  For *Larix* we see clear shifts in climate space when vegetation is held constant (from solid to dashed lines), regardless of whether modern or pre-settlement vegetation distributions are used, indicating that historical climate change plays a strong role in shifting the distribution of *Larix* in climate space across all climate variables.  Historical land use contributes to shifting distributions, particularly for $T_{max}$, where losses at the highest $T_{max}$ values are visible from pre-settlement to modern (Fig. 2d). However, for *Larix*, land use overall appears to exert a weaker influence than climate change.

@@

`r fig_nums("hellPlot", "The total shift in climatic distribution over the last 100 to 150 years for each tree genus ($d_{tot}$) plotted against the amount of shift caused by historic climate change vs. changes in tree distributions ($d_{C} - d_{V}$), the latter mainly due to historic land-use and pathogen outbreaks.  Results are plotted separately for each of the four climate variables considered here. Shifts in distributions are calculated as the Hellinger distance between tree-climate distributions for different combinations of pre-settlement and modern vegetation and climate (Fig. 2).  For $d_{C}-d_{V}$ (Y-axis), negative values indicate a stronger influence of historic land-use on shifting climate distributions, while positive values indicate a stronger influence of historic climate change on shifting distributions.")`

Both the total shift in realized tree-climate distributions and the attribution of these changes to historic land use or climate change varies strongly among taxa and among climate variables (`r fig_nums("hellPlot", display = "cite")`). Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* ($T_{min}$, $T_{max}$), *Larix* ($T_{max}$), *Tsuga* ($T_{max}$ and $P_{ann}$), and *Fagus* ($P_{ann}$).   Attribution analyses based on $d$ (`r fig_nums("hellPlot", display = "cite")`) reveal that while the imprint of historic land use on shifting climatic niches is important for many taxa, this shift is not uniform across taxa. The climatic signal is strongest for $P_{ann}$, with regional changes in $P_{ann}$ being largely responsible for shifting climate niches, while only *Populus* and *Picea* have seen climate niche shifts due to land-use that are larger than climate shifts. The greatest land use signal across taxa is found in $T_{max}$ (`r fig_nums("hellPlot", display = "cite")`), with effects particularly strong for *Larix*, *Ulmus*, and *Fagus* (all taxa lie below the equality line for $d_{c} - d_{v}$). The large shift in the distribution of *Tsuga* relative to $T_{max}$, conversely, appears to be mainly attributable to historic climate change. For $T_{min}$ (`r fig_nums("hellPlot", display = "cite")`), shifting distributions of *Ulmus* and *Thuja*/*Juniperus* can be mainly attributed to land use change, while *Larix* and *Betula* have large shifts that are equally attributable to historic climate change and land use, and the niche shifts for *Quercus*, *Fraxinus*, *Picea*, *Acer*, and *Tilia* are attributable to changing climate.

@@

Compounding and confounding interactions between historic climate change and land use, and the consequent shifts in realized climate space have important implications for the management and genetics of forest communities.  On the one hand, the observation that historical realized niches are broader than modern niches suggests that some tree genera may have more resilience to climate change than expected based on contemporary observations alone. Regionally, annual temperature is expected to increase in Wisconsin by 3.6^o^C in 2055 from a 1980s baseline [@wicci2011climate] and $P_{ann}$ projections range from -17 to 110mm by 2055 relative to 1980s baselines, based on fourteen global climate models from the Climate Model Intercomparison Project [@wicci2011climate]. Every tree genera examined in this study had historical $T_{max}$ values from 0.4 to 1.3^o^C higher than modern estimates, and $P_{ann}$ from 42 to 86mm lower than modern estimates. These higher temperatures and lower precipitation across taxa distributions during the pre-settlement period indicate the potential for greater resilience to climate change than might be assessed using modern data only.  This evidence for historical resilience to warmer and dryer climates may be particularly critical for species that reach their northern or western range limits in the upper Midwest (e.g., *Tsuga*, *Fagus*).

>>>>

# Effect of historical land-use and climate change on tree-climate relationships in the northern United States

@@

Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions is poorly understood. Here we show that tree-climate relationships for 15 common tree genera in the northern US have significantly altered over the last two centuries due to historical land use, pathogen outbreaks, and climate change. Realized niches have shifted towards higher minimum temperatures, cooler maximum temperatures, and higher rainfall. A new attribution method implicates both historical climate change and land use in these shifts, with the relative importance varying among genera and climate variables. Most climate/land-use interactions are compounding, in which historical land-use reinforces climatic shifts in species distributions toward cooler or wetter distributions. The prevalence of compounding interactions imply that contemporary-based models of species distributions may underestimate species resilience to climate change.

@@

Historical land use has substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree] and may cause species distributions to incompletely fill available climates, or otherwise be at disequilibrium [@early2014climatic;@sarmento2013impacts;@ivory2016niche]. Anthropogenic land use is structured and is often biased within climate space, hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature]. Biasing can confound studies that use contemporary distributional data to understand the environmental controls on species distributions and predict species responses to future climate scenarios [@pyke2004habitat].

@@

Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historical climate change, land use, and pathogen outbreaks, and we diagnose the relative influence of historical climate and land use changes on these shifts.  We use gridded historical and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.  We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.  We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and attribute the relative importance of historical vegetation and climate change. We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts. We use this framework to discuss the potential impacts of historical land-use and climate change on inferences about species resilience to climate change, as predicted by species-distribution models based soley on modern distributional data.

@@

`r fig_nums("fourPanel", "(a) The 2x2 factorial design used to diagnose the effects of historical climate and vegetation change on observed shifts in tree-climate relationships. Observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row). (b-e) Representative normalized probability densities for *Larix* for each of the four combinations of climate and vegetation states, plotted relative to $T_{max}$ (b), $T_{diff}$ (c), $T_{min}$ (d), and $P_{ann}$ (e).")`

@@

For most tree genera, historical range losses dominate over range gains and are spatially structured (Table 1, Figure 3). Range losses in the upper Midwest are concentrated in the south and west, along the prairie-savanna-forest ecotone (Figure 3). For *Ulmus* and *Larix*, however, losses occur throughout the range.  Losses are highest for *Ostrya*/*Carpinus* and *Larix* when assessed as a proportion of total range (Supplemental Table 1), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the Upper Midwest.  Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed for *Populus* (`r fig_nums("gain_loss", display="cite")`) and averages `r round(mean(loss_table$Gain),1)`% across all taxa.   Spatial patterns of range gains vary widely among tree genera, and often show less spatial bias than losses (Figure 3).  Range gains for *Picea*, *Larix*, *Abies*, *Tsuga*, *Fagus*, *Ostrya*/*Carpinus*, *Tilia*, and *Fraxinus*, for example, all show a dispersed pattern of gains in a relatively few and widely scattered cells, suggesting that these apparent range gains are mainly caused by sampling uncertainty inherent in local-scale plots and stand heterogeneity, rather than systematic range gains.  Conversely, gains for *Populus* and *Acer*, are concentrated in northern Minnesota, Wisconsin, and Michigan, suggesting range infilling, while *Quercus* gains are consistent with some range extension northwards.  *Pinus*, *Juniperus*/*Thuja*, and *Ulmus* range gains are limited but primarily in the southern portion of their historical range, or beyond it.

@@

The joint effects of historical climate change and land use on tree-climate distributions are illustrated for *Larix* (Figure 2).  For *Larix* we see clear shifts in climate space when vegetation is held constant (from solid to dashed lines), regardless of whether modern or pre-settlement vegetation distributions are used, indicating that historical climate change plays a strong role in shifting the distribution of *Larix* in climate space across all climate variables.  Historical land use contributes to shifting distributions, particularly for $T_{max}$, where losses at the highest $T_{max}$ values are visible from pre-settlement to modern (Fig. 2d). However, for *Larix*, land use overall appears to exert a weaker influence than climate change.

@@

`r fig_nums("hellPlot", "The total shift in climatic distribution over the last 100 to 150 years for each tree genus ($d_{tot}$) plotted against the amount of shift caused by historical climate change vs. changes in tree distributions ($d_{C} - d_{V}$), the latter mainly due to historical land-use and pathogen outbreaks.  Results are plotted separately for each of the four climate variables considered here. Shifts in distributions are calculated as the Hellinger distance between tree-climate distributions for different combinations of pre-settlement and modern vegetation and climate (Fig. 2).  For $d_{C}-d_{V}$ (Y-axis), negative values indicate a stronger influence of historical land-use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.")`

Both the total shift in realized tree-climate distributions and the attribution of these changes to historical land use or climate change varies strongly among taxa and among climate variables (`r fig_nums("hellPlot", display = "cite")`). Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* ($T_{min}$, $T_{max}$), *Larix* ($T_{max}$), *Tsuga* ($T_{max}$ and $P_{ann}$), and *Fagus* ($P_{ann}$).   Attribution analyses based on $d$ (`r fig_nums("hellPlot", display = "cite")`) reveal that while the imprint of historical land use on shifting climatic niches is important for many taxa, this shift is not uniform across taxa. The climatic signal is strongest for $P_{ann}$, with regional changes in $P_{ann}$ being largely responsible for shifting climate niches, while only *Populus* and *Picea* have seen climate niche shifts due to land-use that are larger than climate shifts. The greatest land use signal across taxa is found in $T_{max}$ (`r fig_nums("hellPlot", display = "cite")`), with effects particularly strong for *Larix*, *Ulmus*, and *Fagus* (all taxa lie below the equality line for $d_{c} - d_{v}$). The large shift in the distribution of *Tsuga* relative to $T_{max}$, conversely, appears to be mainly attributable to historical climate change. For $T_{min}$ (`r fig_nums("hellPlot", display = "cite")`), shifting distributions of *Ulmus* and *Thuja*/*Juniperus* can be mainly attributed to land use change, while *Larix* and *Betula* have large shifts that are equally attributable to historical climate change and land use, and the niche shifts for *Quercus*, *Fraxinus*, *Picea*, *Acer*, and *Tilia* are attributable to changing climate.

@@

Compounding and confounding interactions between historical climate change and land use, and the consequent shifts in realized climate space have important implications for the management and genetics of forest communities.  On the one hand, the observation that historical realized niches are broader than modern niches suggests that some tree genera may have more resilience to climate change than expected based on contemporary observations alone. Regionally, annual temperature is expected to increase in Wisconsin by 3.6^o^C in 2055 from a 1980s baseline [@wicci2011climate] and $P_{ann}$ projections range from -17 to 110mm by 2055 relative to 1980s baselines, based on fourteen global climate models from the Climate Model Intercomparison Project [@wicci2011climate]. Every tree genera examined in this study had historical $T_{max}$ values from 0.4 to 1.3^o^C higher than modern estimates, and $P_{ann}$ from 42 to 86mm lower than modern estimates. These higher temperatures and lower precipitation across taxa distributions during the pre-settlement period indicate the potential for greater resilience to climate change than might be assessed using modern data only.  This evidence for historical resilience to warmer and dryer climates may be particularly critical for species that reach their northern or western range limits in the upper Midwest (e.g., *Tsuga*, *Fagus*).
